---
title: "India Urban Climate Analysis 2020–2025"
author: "Brian Kipchumba"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
---

# Business Understanding

India's diverse topography means weather patterns vary drastically across 
cities — from the snow-capped hills of Shimla to the tropical coast of Chennai. 
Understanding these climate patterns is critical for urban planning, agriculture, 
disaster preparedness, and environmental policy.

**Objectives:**

- Explore climate trends across 98 Indian cities from 2020 to 2025
- Analyze seasonal patterns in temperature, rainfall, and wind speed
- Compare climate conditions across geographic regions
- Identify anomalies and extreme weather events
- Forecast future temperature trends using time series models

---

# Data Understanding

## Load Libraries

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(dplyr)
library(readr)
library(scales)
library(viridis)
library(corrplot)
library(forecast)
library(tseries)
library(gridExtra)
library(ggthemes)
```

## Load Dataset

```{r load-data, message=FALSE, warning=FALSE}
df <- read_csv("Data/popular_cities_weather.csv")
cat("Rows:", nrow(df), "\n")
cat("Columns:", ncol(df), "\n")
cat("Cities:", n_distinct(df$city), "\n")
cat("Date Range:", as.character(min(df$date)),
    "to", as.character(max(df$date)), "\n")
```

## Preview Data
```{r preview, message=FALSE, warning=FALSE}
head(df, 10)
```
## Data Structure
```{r structure, message=FALSE, warning=FALSE}
glimpse(df)
```

## Summary Statistics

```{r summary-stats, message=FALSE, warning=FALSE}
summary(df)
```
## 2.6 Missing Values

```{r missing, message=FALSE, warning=FALSE}
missing_vals <- df %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(),
               names_to = "Column",
               values_to = "Missing") %>%
  mutate(Percentage = round(Missing / nrow(df) * 100, 2))

print(missing_vals)

ggplot(missing_vals, aes(x = reorder(Column, -Percentage), y = Percentage)) +
  geom_bar(stat = "identity", fill = "#EF5350") +
  geom_text(aes(label = paste0(Percentage, "%")),
            vjust = -0.5, size = 3.5) +
  labs(title = "Missing Values by Column (%)",
       x = "Column", y = "Missing (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Data Preparation

## Fix Date & Add Features

```{r data-prep, message=FALSE, warning=FALSE}
df <- df %>%
  mutate(
    date      = as.Date(date),
    year      = year(date),
    month     = month(date, label = TRUE),
    month_num = month(date),
    season    = case_when(
      month_num %in% c(12, 1, 2)  ~ "Winter",
      month_num %in% c(3, 4, 5)   ~ "Summer",
      month_num %in% c(6, 7, 8)   ~ "Monsoon",
      month_num %in% c(9, 10, 11) ~ "Post-Monsoon"
    )
  )

cat("Season distribution:\n")
print(table(df$season))
cat("\nNew shape:", nrow(df), "x", ncol(df), "\n")
```

## Handle Missing Values

```{r handle-missing, message=FALSE, warning=FALSE}
# Drop wspd column (100% missing)
df <- df %>% select(-wspd)

# Fill remaining missing values with city mean
df <- df %>%
  group_by(city) %>%
  mutate(
    tavg = ifelse(is.na(tavg), mean(tavg, na.rm = TRUE), tavg),
    tmin = ifelse(is.na(tmin), mean(tmin, na.rm = TRUE), tmin),
    tmax = ifelse(is.na(tmax), mean(tmax, na.rm = TRUE), tmax),
    prcp = ifelse(is.na(prcp), 0, prcp),
    pres = ifelse(is.na(pres), mean(pres, na.rm = TRUE), pres),
    tsun = ifelse(is.na(tsun), mean(tsun, na.rm = TRUE), tsun)
  ) %>%
  ungroup()

cat("Missing values after cleaning:\n")
print(colSums(is.na(df)))
cat("\nFinal shape:", nrow(df), "x", ncol(df), "\n")
cat("Data preparation complete!\n")
```
## Fix Remaining tsun Missing Values

```{r fix-tsun, message=FALSE, warning=FALSE}
# Replace remaining tsun NAs with overall mean
df <- df %>%
  mutate(tsun = ifelse(is.na(tsun), mean(tsun, na.rm = TRUE), tsun))

cat("Missing values after final cleaning:\n")
print(colSums(is.na(df)))
cat("\nFinal shape:", nrow(df), "x", ncol(df), "\n")
```
---

# Temperature Analysis

## National Average Temperature Trend

```{r temp-trend, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
temp_trend <- df %>%
  group_by(date) %>%
  summarise(avg_temp = mean(tavg, na.rm = TRUE), .groups = "drop")

ggplot(temp_trend, aes(x = date, y = avg_temp)) +
  geom_line(color = "#EF5350", linewidth = 1) +
  geom_smooth(method = "loess", se = TRUE, 
              color = "#1565C0", fill = "#BBDEFB") +
  labs(title = "National Average Temperature Trend (2020–2025)",
       x = "Date", y = "Average Temperature (°C)") +
  theme_minimal() +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
The chart shows clear seasonal temperature cycles across India from 2020 to 2025.
Temperatures peak during **Summer (March–May)** reaching up to 32°C and drop 
during **Winter (December–February)** to around 20°C. The smooth blue trend line 
shows a slight overall warming pattern over the 5-year period.

## Monthly Temperature Distribution

```{r monthly-temp, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
ggplot(df, aes(x = month, y = tavg, fill = month)) +
  geom_boxplot(show.legend = FALSE, outlier.alpha = 0.3) +
  scale_fill_viridis_d(option = "C") +
  labs(title = "Monthly Temperature Distribution Across All Cities",
       x = "Month", y = "Average Temperature (°C)") +
  theme_minimal()
```
Temperatures rise steadily from **January** peaking in **May–July** around 30°C 
across all cities. **December and January** show the widest spread, reflecting 
the contrast between cold northern cities like Shimla and warm southern 
cities like Chennai during winter.

## Top 10 Hottest & Coldest Cities

```{r hot-cold, message=FALSE, warning=FALSE, fig.width=12, fig.height=6}
city_temp <- df %>%
  group_by(city) %>%
  summarise(avg_temp = mean(tavg, na.rm = TRUE), .groups = "drop")

top_hot  <- city_temp %>% slice_max(avg_temp, n = 10)
top_cold <- city_temp %>% slice_min(avg_temp, n = 10)

p1 <- ggplot(top_hot, aes(x = reorder(city, avg_temp), y = avg_temp)) +
  geom_bar(stat = "identity", fill = "#EF5350") +
  geom_text(aes(label = round(avg_temp, 1)), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "Top 10 Hottest Cities", x = "", y = "Avg Temp (°C)") +
  theme_minimal()

p2 <- ggplot(top_cold, aes(x = reorder(city, -avg_temp), y = avg_temp)) +
  geom_bar(stat = "identity", fill = "#42A5F5") +
  geom_text(aes(label = round(avg_temp, 1)), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "Top 10 Coldest Cities", x = "", y = "Avg Temp (°C)") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```
**Hottest cities** are concentrated in South India — coastal and inland cities 
like Tiruchirappalli and Kozhikode maintain high temperatures year-round due 
to their tropical location. **Coldest cities** are in North India — Srinagar 
and Shimla are hill stations with significantly lower average temperatures 
due to high altitude.

##  Seasonal Temperature Analysis

```{r seasonal-temp, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
df %>%
  group_by(season) %>%
  summarise(
    Average = mean(tavg, na.rm = TRUE),
    Minimum = mean(tmin, na.rm = TRUE),
    Maximum = mean(tmax, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(Average, Minimum, Maximum),
               names_to = "type", values_to = "temp") %>%
  ggplot(aes(x = season, y = temp, fill = type)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = round(temp, 1)),
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("#42A5F5", "#66BB6A", "#EF5350")) +
  labs(title = "Average Temperature by Season",
       x = "Season", y = "Temperature (°C)", fill = "Type") +
  theme_minimal()
```
**Summer** records the highest maximum temperatures (35.2°C) while **Winter** 
has the lowest minimum (14.9°C). **Monsoon** maintains warm average temperatures 
despite heavy rainfall. The temperature range is widest in Winter reflecting 
the contrast between northern and southern cities.
---

#  Rainfall Analysis

## Monthly Rainfall Pattern

```{r rainfall-monthly, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
df %>%
  group_by(month, month_num) %>%
  summarise(avg_rain = mean(prcp, na.rm = TRUE), .groups = "drop") %>%
  arrange(month_num) %>%
  ggplot(aes(x = month, y = avg_rain, fill = avg_rain)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_c(option = "B") +
  labs(title = "Average Monthly Rainfall Across All Cities",
       x = "Month", y = "Average Rainfall (mm)", fill = "Rain (mm)") +
  theme_minimal()
```
Rainfall is heavily concentrated in the **Monsoon months (June–September)** 
with July recording the highest average of ~320mm across all cities. 
The dry season from **January to April** shows near-zero rainfall, 
confirming India's strong monsoon-driven precipitation pattern.

## Top 10 Wettest Cities

```{r wettest-cities, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
df %>%
  group_by(city) %>%
  summarise(total_rain = sum(prcp, na.rm = TRUE), .groups = "drop") %>%
  slice_max(total_rain, n = 10) %>%
  ggplot(aes(x = reorder(city, total_rain), y = total_rain)) +
  geom_bar(stat = "identity", fill = "#1565C0") +
  geom_text(aes(label = round(total_rain, 0)), hjust = -0.2, size = 3) +
  coord_flip() +
  labs(title = "Top 10 Wettest Cities (Total Rainfall 2020–2025)",
       x = "", y = "Total Rainfall (mm)") +
  theme_minimal()
```
**Mangalore** is the wettest city with over 25,000mm total rainfall over 
5 years, followed by **Kozhikode** and several Mumbai metropolitan cities. 
All top 10 wettest cities are located on India's **western coastal belt**, 
which receives the heaviest Southwest Monsoon rainfall.

## Rainfall by Season

```{r rainfall-season, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
df %>%
  group_by(season) %>%
  summarise(avg_rain = mean(prcp, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = reorder(season, -avg_rain), y = avg_rain, fill = season)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = round(avg_rain, 1)), vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("#1565C0", "#EF5350", "#66BB6A", "#FFA726")) +
  labs(title = "Average Rainfall by Season",
       x = "Season", y = "Average Rainfall (mm)") +
  theme_minimal()
```
**Monsoon season dominates** with 239.7mm average rainfall — more than 
double Post-Monsoon. Winter is the driest season with only 17.4mm, 
confirming that over **58% of India's annual rainfall** comes during 
the June–September Monsoon period.

---

# Pressure & Correlation Analysis

##  Average Pressure by Month

```{r pressure, message=FALSE, warning=FALSE, fig.width=10, fig.height=5}
df %>%
  group_by(month, month_num) %>%
  summarise(avg_pres = mean(pres, na.rm = TRUE), .groups = "drop") %>%
  arrange(month_num) %>%
  ggplot(aes(x = month, y = avg_pres, group = 1)) +
  geom_line(color = "#AB47BC", linewidth = 1.2) +
  geom_point(color = "#AB47BC", size = 3) +
  labs(title = "Average Atmospheric Pressure by Month",
       x = "Month", y = "Pressure (hPa)") +
  theme_minimal()
```
Atmospheric pressure follows an **inverse pattern to rainfall** — pressure 
drops sharply from January to June/July coinciding exactly with the arrival 
of the Monsoon. The lowest pressure in **June–July (1002 hPa)** corresponds 
to the peak rainfall months, confirming that low pressure systems drive 
India's Monsoon precipitation.

## 6.2 Correlation Matrix

```{r correlation, message=FALSE, warning=FALSE, fig.width=8, fig.height=7}
cor_data <- df %>%
  select(tavg, tmin, tmax, prcp, pres, tsun) %>%
  drop_na()

cor_matrix <- cor(cor_data)

corrplot(cor_matrix,
         method      = "color",
         type        = "upper",
         addCoef.col = "black",
         tl.col      = "black",
         tl.srt      = 45,
         title       = "Correlation Matrix of Climate Variables",
         mar         = c(0, 0, 2, 0))
```
Temperature variables are **strongly correlated** with each other (0.94–0.95). 
The strongest negative correlation is between **minimum temperature and pressure 
(-0.83)** — as temperatures rise, pressure drops driving Monsoon rainfall. 
Rainfall and sunshine show a **negative correlation (-0.39)** as expected — 
rainy months have significantly less sunshine.

---

#  City Comparison

##  Temperature Trends — Major Cities

```{r city-comparison, message=FALSE, warning=FALSE, fig.width=12, fig.height=6}
major_cities <- c("Mumbai", "Delhi", "Chennai", 
                  "Kolkata", "Bangalore", "Hyderabad")

df %>%
  filter(city %in% major_cities) %>%
  group_by(city, date) %>%
  summarise(avg_temp = mean(tavg, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = date, y = avg_temp, color = city)) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  scale_color_viridis_d() +
  labs(title = "Temperature Trends — Major Indian Cities (2020–2025)",
       x = "Date", y = "Average Temperature (°C)", color = "City") +
  theme_minimal() +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
**Delhi** shows the most extreme seasonal variation — dropping to 14°C in 
winter and rising to 34°C in summer. **Mumbai and Chennai** maintain stable 
temperatures year-round due to their coastal location. **Bangalore** stays 
the coolest among southern cities due to its high elevation.

# Stationarity Test — Mumbai

```{r stationarity, message=FALSE, warning=FALSE}
# Extract Mumbai temperature time series
mumbai_ts <- df %>%
  filter(city == "Mumbai") %>%
  arrange(date) %>%
  pull(tavg)

mumbai_ts <- ts(mumbai_ts, start = c(2020, 1), frequency = 12)

# Plot the time series first
autoplot(mumbai_ts) +
  labs(title = "Mumbai Temperature Time Series",
       x = "Year", y = "Temperature (°C)") +
  theme_minimal()
```
Clear seasonal cycles every year
No upward/downward drift — confirms stationarity

##  ADF Test (Augmented Dickey-Fuller)

```{r adf-test, message=FALSE, warning=FALSE}
# ADF Test — tests for stationarity
# H0: Series is NON-stationary
# H1: Series IS stationary
adf_result <- adf.test(mumbai_ts)
print(adf_result)

cat("\n")
if (adf_result$p.value < 0.05) {
  cat("✅ p-value =", round(adf_result$p.value, 4), "\n")
  cat("✅ Series is STATIONARY — no differencing needed\n")
} else {
  cat("❌ p-value =", round(adf_result$p.value, 4), "\n")
  cat("❌ Series is NON-STATIONARY — differencing needed\n")
}
```
p-value = 0.0243 (less than 0.05)
Series is STATIONARY — no differencing needed


##  ACF & PACF Plots

```{r acf-pacf, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
# ACF and PACF help determine ARIMA parameters
par(mfrow = c(1, 2))
acf(mumbai_ts,  main = "ACF — Mumbai Temperature")
pacf(mumbai_ts, main = "PACF — Mumbai Temperature")
```
ACF shows significant spikes at lag 1.0 — strong seasonality
PACF cuts off quickly — suggests AR terms needed

##  ARIMA Model — Mumbai

```{r arima-mumbai, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
# auto.arima automatically selects best ARIMA parameters
fit_mumbai <- auto.arima(mumbai_ts, 
                          seasonal = TRUE,
                          stepwise = FALSE,
                          approximation = FALSE)

cat("Best ARIMA Model for Mumbai:\n")
summary(fit_mumbai)
```

##  Forecast — Next 12 Months

```{r forecast-mumbai, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
forecast_mumbai <- forecast(fit_mumbai, h = 12)

autoplot(forecast_mumbai) +
  labs(title = "Mumbai Temperature Forecast — Next 12 Months (2026)",
       x = "Year", y = "Temperature (°C)") +
  theme_minimal() +
  theme(plot.title = element_text(fontface = "bold"))
```

##  Model Accuracy

```{r accuracy-mumbai, message=FALSE, warning=FALSE}
acc <- accuracy(fit_mumbai)
cat("Model Accuracy Metrics:\n")
print(acc)

cat("\nForecast Values for Next 12 Months:\n")
print(round(forecast_mumbai$mean, 2))
```
The **ARIMA(0,0,0)(2,1,0)[12]** model captures Mumbai's seasonal temperature 
pattern with high accuracy — **MAPE of only 1.84%** meaning predictions are 
within 2% of actual values. The 2026 forecast shows peak temperatures of 
**30.5°C in May** dropping to **25.4°C in January**, consistent with 
Mumbai's historical seasonal pattern.

# Stationarity Test — Delhi

```{r delhi-stationarity, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
delhi_ts <- df %>%
  filter(city == "Delhi") %>%
  arrange(date) %>%
  pull(tavg)

delhi_ts <- ts(delhi_ts, start = c(2020, 1), frequency = 12)

# Plot Delhi time series
autoplot(delhi_ts) +
  labs(title = "Delhi Temperature Time Series",
       x = "Year", y = "Temperature (°C)") +
  theme_minimal()
```

## Delhi ADF Test

```{r delhi-adf, message=FALSE, warning=FALSE}
adf_delhi <- adf.test(delhi_ts)
print(adf_delhi)

cat("\n")
if (adf_delhi$p.value < 0.05) {
  cat("✅ p-value =", round(adf_delhi$p.value, 4), "\n")
  cat("✅ Delhi series is STATIONARY — no differencing needed\n")
} else {
  cat("❌ p-value =", round(adf_delhi$p.value, 4), "\n")
  cat("❌ Delhi series is NON-STATIONARY — differencing needed\n")
}
```

## 8.8 Delhi ACF & PACF

```{r delhi-acf, message=FALSE, warning=FALSE}
par(mfrow = c(1, 2))
acf(delhi_ts,  main = "ACF — Delhi Temperature")
pacf(delhi_ts, main = "PACF — Delhi Temperature")
```

## ARIMA Model — Delhi

```{r arima-delhi, message=FALSE, warning=FALSE}
fit_delhi <- auto.arima(delhi_ts,
                         seasonal = TRUE,
                         stepwise = FALSE,
                         approximation = FALSE)

cat("Best ARIMA Model for Delhi:\n")
summary(fit_delhi)
```

##  Delhi Forecast — Next 12 Months

```{r forecast-delhi, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
forecast_delhi <- forecast(fit_delhi, h = 12)

autoplot(forecast_delhi) +
  labs(title = "Delhi Temperature Forecast — Next 12 Months (2026)",
       x = "Year", y = "Temperature (°C)") +
  theme_minimal() +
  theme(plot.title = element_text(fontface = "bold"))
```
The **Delhi ARIMA(0,0,1)(0,1,2)[12]** model forecasts a peak temperature 
of **33°C in June 2026** dropping sharply to **16°C in December 2026**. 
The narrow confidence interval (blue shading) indicates high forecast 
confidence. Delhi's extreme seasonal swing of nearly **20°C** between 
winter and summer is clearly captured by the model.

## Mumbai vs Delhi Forecast Comparison

```{r compare-forecast, message=FALSE, warning=FALSE, fig.width=12, fig.height=5}
mumbai_fc <- as.numeric(forecast_mumbai$mean)
delhi_fc  <- as.numeric(forecast_delhi$mean)
months    <- c("Jan","Feb","Mar","Apr","May","Jun",
               "Jul","Aug","Sep","Oct","Nov","Dec")

forecast_compare <- data.frame(
  Month  = factor(months, levels = months),
  Mumbai = mumbai_fc,
  Delhi  = delhi_fc
) %>%
  pivot_longer(cols = c(Mumbai, Delhi),
               names_to  = "City",
               values_to = "Temperature")

ggplot(forecast_compare, aes(x = Month, y = Temperature,
                              color = City, group = City)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("#EF5350", "#1565C0")) +
  labs(title = "2026 Temperature Forecast — Mumbai vs Delhi",
       x = "Month", y = "Forecasted Temperature (°C)") +
  theme_minimal() +
  theme(plot.title = element_text(fontface = "bold"))
```
The comparison chart highlights the fundamental climate difference between 
the two cities. **Delhi (red)** shows extreme seasonal variation — starting 
at 13°C in January, peaking at 33°C in June, then dropping back to 16°C 
in December. **Mumbai (blue)** remains remarkably stable throughout the year, 
staying between 25°C and 30°C — never too cold, never extreme. This reflects 
the difference between Delhi's **continental inland climate** and Mumbai's 
**tropical coastal climate** moderated by the Arabian Sea.

---

# Conclusion & Recommendations

##  Key Findings

This project analyzed monthly climate data for 96 Indian cities from 
2020 to 2025, covering temperature, rainfall, pressure, and sunshine duration.

**Temperature:**

- India shows strong seasonal temperature cycles with peak temperatures 
  in **May–June (Summer)** averaging 28.6°C and lowest in 
  **December–January (Winter)** averaging 20.6°C
- **Tiruchirappalli and Kozhikode** are the hottest cities (~29°C average)
- **Srinagar and Shimla** are the coldest cities (14–16°C average)
- **Delhi** shows the most extreme seasonal variation — 21°C range between 
  winter and summer

**Rainfall:**

- **Monsoon season (June–September)** dominates rainfall with 239.7mm 
  average — more than double all other seasons combined
- **Mangalore** is the wettest city with over 25,000mm total rainfall 
  over 5 years
- All top 10 wettest cities are located on India's western coastal belt

**Pressure & Correlation:**

- Atmospheric pressure has a **strong inverse relationship with temperature** 
  (−0.79) and rainfall (−0.41)
- Pressure drops to its lowest in **June–July** coinciding exactly with 
  peak Monsoon rainfall
- Rainfall and sunshine are **negatively correlated (−0.39)** — rainy 
  months have significantly less sunshine

**Forecasting:**

- Both Mumbai and Delhi temperature series are **stationary** — confirmed 
  by ADF test (p < 0.05)
- **Mumbai ARIMA(0,0,0)(2,1,0)[12]** — MAPE of 1.84% — highly accurate
- **Delhi ARIMA(0,0,1)(0,1,2)[12]** — MAPE of 2.69% — highly accurate
- 2026 forecasts show temperatures will follow historical seasonal patterns

##  Recommendations

- **Urban Planning:** Cities with high summer temperatures like Tiruchirappalli 
  should invest in green cover, water bodies, and heat-resilient infrastructure
- **Agriculture:** Farmers in rain-dependent regions should use these forecasts 
  for crop planning especially during irregular Monsoon years
- **Disaster Preparedness:** Coastal cities like Mumbai, Mangalore, and 
  Kozhikode should strengthen flood management systems given consistently 
  high rainfall
- **Climate Monitoring:** The slight warming trend observed in the national 
  average temperature warrants closer environmental monitoring and 
  policy intervention



*Analysis completed using R 4.5.0 | 
Data: Indian Cities Weather Dataset 2020–2025*
